---
title: "final_cleaning"
output: html_document
date: "2024-12-27"
---

### **Loading data**

```{r}
getwd()
setwd("~/pathway/to/Group5")

# loading IMPC data into files 
files <- list.files(path = '5', pattern='csv', full.names = TRUE)
print(files[1:3]) #prints first three files

sample_dat <- lapply((files), function(x){
  read.csv(x, header=FALSE, sep=",", row.names = 1)})
View(sample_dat)


print(sample_dat [1:3])
#NB: When working with larger files (e.g., when we do the whole dataset) can use the fread() function from the data.table function instead

```

### **Ordering the row names**

```{r}
sample_dat_ordered <- lapply((sample_dat), function(y){
  y[order(row.names(y)), , drop=FALSE]
}) #Sort rows by their row names and orders them so they are all presented in the same order 

#drop = FALSE prevents them from being dropped/removed - losing dimension as a dataframe
#NB: did not check for missing rownames or duplicated rownames
```

```{r}
print(sample_dat_ordered[1:3])
View(sample_dat_ordered)

```

### **Rearranging the data using dplyr**

```{r}
install.packages("dplyr")
library(dplyr)

sample_dat_ordered <- as.data.frame(sample_dat_ordered) #convert into dataframe to make it quicker and more efficient to bind
new_sample_dat <- bind_cols(sample_dat_ordered)
new_sample_dat <- t(new_sample_dat)
View(new_sample_dat)
typeof(new_sample_dat)


is.data.frame(new_sample_dat) #FALSE
is.list(new_sample_dat) #FALSE 

#Currently character not dataframe so convert into dataframe 
new_sample_dat <- as.data.frame(new_sample_dat)
is.data.frame(new_sample_dat) #TRUE

new_sample_dat <- as.data.frame(lapply(new_sample_dat, type.convert, as.is = TRUE)) #Converts all the data in each column into the right datatypes
View(new_sample_dat)

```

### Data Cleaning - Checking for NAs

```{r}

#Check for NA's in all columns 
for (n in colnames(new_sample_dat)){
  print(sum(is.na(new_sample_dat[[n]])))
}
# 0 across all columns - No NA's found 
```

```{r echo=TRUE}
#First checking mouse_life_stages
print(unique(new_sample_dat[4])) #checking all inputted life stages
```

No abnormal life stages - no need for further cleaning

### Data Cleaning - Checking for outliers

#### **Step 1: Checking mouse_strain**

```{r}
print(unique(new_sample_dat[5]))
print(unique(new_sample_dat[5] == 'C3H')) #[1] FALSE - there are no C3H strains
#21 unique values; should be 3 (excluding C3H) - need to remove 18

sub_dat <- subset(new_sample_dat,mouse_strain %in% c('C57BL','129SV', 'C3H','B6J' ))

print(unique(sub_dat[5])) #Now we only have data files with either C57Bl, 129SV or B6J
View(sub_dat)
```

``` r
Description:df [21 × 1]
 
 
mouse_strain
<chr>
1   C57BL           
2   C54BL           
5   C50BL           
108 C58BL           
137 129SV           
215 B6J         
291 C53BL           
349 C56BL           
1271    C52BL           
1713    C51BL           
1-10 of 21 rows
```

#### **Step 2: Checking p-value**

```{r}
min(sub_dat[8]) #checking the minimum p-value stored [1] 0 -> so it's within range
max(sub_dat[8]) #checking the maximum p-value stored [1] 1.499989 - so we need to restrict it to 1

sub_dat[8] <- mapply(function(x) min(1,x),sub_dat[[8]])
max(sub_dat[8]) # [1] 1 - now  within range

```

#### **Step 3: Checking gene_accession_id**

```{r}
#Check format of gene_accession_id 
#a. length = 9-11
#gene_accession_id is in column 2 
id_column <- sub_dat[[2]]
print(id_column)

# Count rows where the ID length is shorter than 9 or longer than 11
id_length <- (which(nchar(id_column) < 9 | nchar(id_column) > 11))
print(sub_dat[id_length,]) #Nothing printed so they are all the right length 

```

### Data Cleaning - Checking for duplicates

#### **Step 4: Checking for duplicated entries for gene_symbol and gene accession_id**

-   Relationship between gene_accession_id and gene_symbol should be one-to-one

```{r}
print(unique(sub_dat[3])) #[1] 393 unique genes_symbols
print(unique(sub_dat[2])) #[1] 200 unique gene_accession_id

#Multiple gene symbols being used for one gene_accession_id 
#We need to find which gene id's have multiple gene symbols 

id_and_symbol_dat <- unique(sub_dat[2:3]) #paired dataframe containing all unique values of gene symbols and accession ids.
print(id_and_symbol_dat)

id_symbols <- aggregate(id_and_symbol_dat[2],list(gene_id = id_and_symbol_dat$gene_accession_id),function(x) x)
View(id_symbols) #aggregate - group of paired sets

print(unique(mapply(length,id_symbols$gene_symbol))) #[1] 1 2 - for each accession id it either has one gene symbol input or two gene symbol input => there are no more than two genes for each id. 

dup_flag <- mapply(function(x) length(x) > 1,id_symbols$gene_symbol)
dup_genes <- subset(id_symbols,dup_flag)
print(dup_genes)

```

``` r
sub_dat$gene_accession_id sub_dat$gene_symbol
1 MGI:103225               c("Cyp4b1", "CYP4B1")
2 MGI:104629               c("Penk", "PENK")
3 MGI:105304               c("Il6ra", "IL6RA")
4 MGI:106924               c("Melk", "MELK")
5 MGI:107164               c("Ppp3ca", "PPP3CA")
```

-   All duplications are due to differences in entries as lower case and upper case - can remove the uppercase entries and replace with lowercase

```{r}
install.packages("stringr")
library(stringr)
apply_lower_case <- str_to_title(sub_dat$gene_symbol) #str_to_title function from stringr capitalizes the first character and applies lowercase on rest of string
sub_dat$gene_symbol <- apply_lower_case
View(sub_dat)

print(unique(sub_dat[3])) #[1] 200 - matches the number of unique gene accession_ids
```

#### **Step 5: Checking for entirely duplicated entries**

-   Are there any complete duplicated rows (all 8 cols identical values) - FALSE
-   Are there multiple rows for the same phenotype & mouse model (all cols except analysis_id identical) - TRUE
-   Also many experiment repeats for the same mouse model & phenotype (diff analysis_id & pvalue)

**Checking for complete identical entries**

```{r, echo=TRUE}

all_cols <- c("analysis_id","gene_accession_id","gene_symbol","mouse_life_stage","mouse_strain","parameter_id","parameter_name")

complete_duplicates <- sub_dat %>%
  group_by(across(all_of(all_cols))) %>%
  filter(n_distinct(pvalue) > 1) %>% #keep grps with distinct p-values 
  ungroup()

View(complete_duplicates) #no complete duplicates  

print(unique(sub_dat[1])) #[1] 193,098 - no duplicated analysis_id 

```

**Checking for multiple entries for the same mouse model & phenotype**

-   Are the p-values duplicated suggesting duplication
-   Are p-values different suggesting experimental repeats

```{r}

subset_cols <- c("gene_accession_id","gene_symbol","mouse_life_stage","mouse_strain","parameter_id","parameter_name")

check <- sub_dat %>%
  group_by(across(all_of(subset_cols))) %>%  # Group by identical columns
  summarise(count = n(), unique_p_values = n_distinct(pvalue), .groups = "drop") %>% #count no. of repeats & unique pvalues
  filter(unique_p_values > 1) #keeps groups with distinct pvalues

View(check)

# Filter rows where 'count' is not equal to 'unique_p_value'
mismatched_rows <- check %>% 
  filter(count != unique_p_values)
# Print the mismatched rows
View(mismatched_rows)

#identifies all entries where duplications are present, as well as alternative experimental p-values 

```

#### **Step 6: Checking for duplicated p-values of entries with the same column values excluding analysis_id**

**Method One - try and identify the duplicated p-values**

```{r}
print(unique(sub_dat[8])) #[1] 173,305 unique p-values however although they may be unique they may be for the same exact gene, strain,life stage, and parameter -> currently there are 193,098 entries in total so need to remove 19,793 duplicates 
```

``` r
duplicated_p_value <- function(x){
  x <- sub_dat[duplicated(sub_dat$pvalue),]
} 

duplicated_p_value <- sub_dat[duplicated(sub_dat[8]),]
View(duplicated_p_value)
View(unique(duplicated_p_value[8]))
```

This doesn't identify duplicated pvalues of specifically the ones with the same gene id, symbol, strain, life stage, parameter id and name; just duplicated pvalues in the entire dataset so there could be two different entries with the same value - **not selective**

**Method Two**

This method was able to identify all duplicated entries without flagging unique entries with coincidentally repeated pvalues

```{r}
repeated_rows <- sub_dat %>%
  group_by(across(2:8)) %>% #groups all identical cols except analysis_id
  filter(n() > 1) %>% #keep rows that have been repeated 
  ungroup() 
  
View(repeated_rows) #38,046 rows of duplicates


# check all pvalues appear at least twice 
single_occurrence_check <- repeated_rows %>%
  group_by(pvalue) %>%  # Group by pvalue
  summarise(count = n(), .groups = "drop")  # Count occurrences of each pvalue
 
View(single_occurrence_check)

```

**Method Three - (Chosen Method)**

```{r}

library(dplyr) #Required for the group_by function
grouped_gene_id_and_symbol <- sub_dat %>%
  group_by(gene_accession_id, gene_symbol, mouse_life_stage, mouse_strain, parameter_id, parameter_name) %>%
  filter(duplicated(pvalue) | duplicated(pvalue, fromLast = TRUE)) %>% 
  slice(1) %>%
  ungroup()
View(grouped_gene_id_and_symbol) 

#Testing on sub_dat to see if the duplicated rows correlate 
print(subset(sub_dat, sub_dat$gene_symbol== "Ndufs1" & sub_dat$parameter_name == "Total protein")) #prints both duplicates

print(subset(grouped_gene_id_and_symbol, grouped_gene_id_and_symbol$gene_symbol== "Ndufs1" & grouped_gene_id_and_symbol$parameter_name == "Total protein")) #prints only the one record it retained which is the first duplicate it came across with the analysis_id 00096jlj53g968z

```

There is an entry in the grouped_gene_id_symbol for the Ndufs1 gene_symbol where parameter_name was Total protein - there were no other duplicates found for Ndufs1 so this is the only duplicate associated with this gene. Checked sub_dat and two entries were printed with the same p-value

-   `slice(1)` : Makes it so it only stores the record including the analysis_id of the first duplicate it comes across

#### **Removing duplicates from the sub_dat dataset**

```{r}
#Using anti_join from dplyr package
sub_dat_2 <- anti_join(sub_dat, grouped_gene_id_and_symbol) #uses the duplicate data frame to only retained values that are NOT in the duplicate dataset
View(sub_dat_2) 
print(subset(sub_dat_2, sub_dat_2$gene_symbol== "Ndufs1" & sub_dat_2$parameter_name == "Total protein")) #Only one copy - all duplicates have been removed

```

### **Writing cleaned dataset as .csv to push into HPC**

```{r}
#saving cleaned data as csv file for push into HPC
write.csv(sub_dat_2,"/Users/path/to/directory/final_cleaned_data.csv", row.names=FALSE, quote=FALSE) 

```
